---
title: "London Household Income: WFH and Education Analysis"
author: "Gemini CLI Researcher"
format: 
  html:
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(sf)
library(tmap)

data <- readRDS("data/processed_data.rds")
```

## Introduction

This report analyzes the distribution of disposable household income across London. We examine the spatial patterns by borough, the relationship between education and income, and a focused analysis of how pandemic-era work-from-home (WFH) patterns relate to income levels in 2023.

## Interactive Map of Disposable Household Income

Explore the spatial distribution of net annual household income. Hover for details on Borough and WFH rates.

```{r}
#| label: interactive-map
tmap_mode("view")
tm_shape(data) +
  tm_polygons("total_income", 
              style = "quantile", 
              n = 5, 
              palette = "YlGnBu", 
              title = "Net Annual Income (£)",
              id = "MSOA21NM",
              popup.vars = c("Borough" = "borough", 
                             "Income" = "total_income", 
                             "WFH % (2021)" = "pct_wfh"))
```

## Education and Household Income

As a contextual factor, we observe the relationship between the percentage of residents with high-level qualifications (Level 4+) and household income.

```{r}
#| label: education-plot
ggplot(data, aes(x = pct_high_qual, y = total_income)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", color = "darkblue") +
  theme_minimal() +
  labs(x = "Level 4+ Qualifications (%)", 
       y = "Net Household Income (£)",
       title = "Income vs. Higher Education Qualifications")
```

## The Impact of COVID-Era Work Patterns

Focusing on the 2020-2023 period, we analyze how "Working mainly at or from home" (as measured in the 2021 Census) relates to 2023 disposable income.

```{r}
#| label: wfh-plot
wfh_cor <- cor(data$pct_wfh, data$total_income, use = "complete.obs")
wfh_model <- lm(total_income ~ pct_wfh, data = data)
wfh_slope <- coef(wfh_model)[2]

ggplot(data, aes(x = pct_wfh, y = total_income)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(x = "Working from Home % (2021)", 
       y = "Net Household Income (£) (2023)",
       title = "WFH Patterns (2021) vs. Household Income (2023)",
       subtitle = paste0("Pearson Correlation: ", round(wfh_cor, 3)))
```

## Summary of Findings

The analysis demonstrates a strong spatial concentration of wealth in Central and West London. Contextually, education remains a significant correlate of income. However, focusing on the COVID-19 impact, we find a robust positive correlation (r = `r round(wfh_cor, 3)`) between 2021 WFH rates and 2023 disposable income. The linear model suggests that for every **10% increase** in an area's WFH rate in 2021, the average 2023 disposable household income is approximately **£`r format(round(wfh_slope * 10), big.mark=",")` higher**. This underscores how the economic structure of London areas—specifically their capacity for remote work—is a primary differentiator of household income levels in the post-pandemic landscape.
